---
title: "HW 3"
output: html_notebook
---


```{r}
ad = read.csv('Advertising_Updated.csv')

lm <- lm(Sales~., data=ad)
new.dat <- data.frame(TV=200, Radio=10, Newspaper=20)

predict(lm, newdata = new.dat, interval = "confidence")
```
# Q.1

```{r}
library(dplyr)
library(magrittr)
data = read.csv('KAG.csv', stringsAsFactors = FALSE)

head(data)



group_by(data, ad_id) %>% 
  summarise(., cpc = sum(CPC), imp = sum(Impressions)) %>% 
  filter(cpc < 1) %>% 
  arrange(., cpc, desc(imp))



```
The ad among ads with cpc = 0 that led to the most impressions is 1121094.

# Q.2
```{r}
group_by(data, campaign_id) %>% 
  summarise(., spent = sum(Spent), imp = sum(Impressions), ) %>% 
  mutate(., cpm=spent/imp) %>%
  arrange(., cpm)
```

Campaign 1178 was the least efficient on brand awareness on average.

# Q.3

```{r}

summarise(data, spent = sum(Spent), tc = 5*sum(CostPerConv_Total), ac = 50*sum(CostPerConv_Approved), ) %>% 
  mutate(., roas=(tc+ac)/spent) %>%
  arrange(., roas)
```
ROAS is 34.32.


```{r}
library(ggplot2)

d = filter(data, interest==15 | interest==21 | interest==101) %>%
  mutate(., interest=factor(interest)) %>%
  group_by(., interest, gender) %>% 
  mutate(., roas=(CostPerConv_Total+CostPerConv_Approved)/Spent) %>%
  filter(!is.na(roas)) %>%
  arrange(., roas)

ggplot(d, aes(x = interest, y = roas, group = interest)) + 
  geom_boxplot() +
  labs(x="Interest ID", y = "ROAS")
```

```{r}
filter(data, campaign_id == 1178) %>%
  mutate(., roas=(CostPerConv_Total+CostPerConv_Approved)/Spent, gender=factor(gender)) %>%
  filter(!is.na(roas)) %>%
  group_by(., gender) %>% 
  summarise(., mn = mean(roas), md = median(roas)) 
  
```

# Q.5
```{r}
library(readr)
library(tidyverse)
library(correlationfunnel)
library(DataExplorer)
library(WVPlots)
library(ggthemes) 
library(ROCR)
library(caret)
library(corrplot)




```

## a)

```{r}
data = read.csv('advertising1.csv', stringsAsFactors = TRUE)

head(data)



ggplot(data, aes( y = Daily.Time.Spent.on.Site, )) + 
  geom_boxplot() +
  labs(y = "Daily.Time.Spent.on.Site")

ggplot(data, aes( y = Daily.Internet.Usage, )) + 
  geom_boxplot() +
  labs(y = "Daily.Internet.Usage")

ggplot(data, aes( y = Area.Income, )) + 
  geom_boxplot() +
  labs(y = "Area.Income")

```
## b)

```{r}
data

ggplot(data=data, aes(x=Male, y=Clicked.on.Ad)) +
  geom_bar(stat="identity")

ggplot(data=data, aes(x=Age, y=Clicked.on.Ad)) +
  geom_bar(stat="identity")
```

## c)

```{r}
ggplot(data, aes( x= factor(Clicked.on.Ad),y = Age, )) + 
  geom_boxplot() +
  labs(y = "Age")


ggplot(data, aes( x= factor(Clicked.on.Ad),y = Area.Income, )) + 
  geom_boxplot() +
  labs(y = "Area.Income")


ggplot(data, aes( x= factor(Clicked.on.Ad),y = Daily.Internet.Usage, )) + 
  geom_boxplot() +
  labs(y = "Daily.Internet.Usage")


ggplot(data, aes( x= factor(Clicked.on.Ad),y = Daily.Time.Spent.on.Site, )) + 
  geom_boxplot() +
  labs(y = "Daily.Time.Spent.on.Site")
```
## d)  

Based on our preliminary boxplots, I would expect an older person to be more likely to click on the ad than someone younger.



# Q.6


```{r}
ggplot(data, aes(x=Age, y=Area.Income, shape=factor(Clicked.on.Ad), color=factor(Clicked.on.Ad))) +
  geom_point()
```



Based on this plot I would not expect a 32 year old making an income of $62,000 to click on the add, because it seems like clicks have a positive correlation with age and negative correlation with income.  The majority of clicks happer at ages > 32 and incomes < $60,000

```{r}
ggplot(data, aes(x=Age, y=Daily.Time.Spent.on.Site, shape=factor(Clicked.on.Ad), color=factor(Clicked.on.Ad))) +
  geom_point()
```

Based on this plot, I would not expect a 50-year-old person who spends 60 minutes daily on the site to click on the ad because there seems to be a negative correlation between time spent on the site and clicks.  People that spend more time on the site are less likely to click on the ad.

## Q.7
```{r}
data %>%
mutate_if(is.numeric,as.numeric)%>%
binarize() %>%
correlate(Clicked.on.Ad__1) %>%
plot_correlation_funnel(interactive = TRUE, alpha = 0.7)
```

### b)
```{r}
model = glm(Clicked.on.Ad~Daily.Internet.Usage + Daily.Time.Spent.on.Site + Age + Area.Income,data=data,family=binomial())
summary(model)
```
## Q.8
```{r}

data$predict = predict(model, data, type="response")>=.8
data$test = data$Clicked.on.Ad==1
confusionMatrix(factor(data$predict), factor(data$test))


```
